---
title: "Project Proposal"
author: "Gifty Osei"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(tidyr)
library(kableExtra)
```


#  Unveiling Temporal Health Patterns: A Bayesian MCMC and Clustering Approach to Health Indicators

```{r, eval=FALSE, include=FALSE, echo=FALSE}
#- Description
#- Questions to answer
#- Data set
#- Statistical Computing Methods
#- Enough details

#- Describe research motivation, data source and structure and justify variables and statistical models to be used.

```





## Motivation

Health disparities between demographics and geographic areas remain a major public health concern. Understanding how different health indicators change over time by gender, race/ethnicity, and geography is critical for making informed policy decisions and effectively directing interventions. The goal of this project is to examine a large dataset comprising a variety of health indicators in order to discover underlying patterns, trends, and clusters. We use advanced statistical computing approaches such as Bayesian modeling with Markov Chain Monte Carlo (MCMC), longitudinal analysis, and clustering techniques to provide insights into the dynamics of health indicators and overcome difficulties such as missing data. The ultimate goal is to allow the data to disclose intrinsic structures and relationships, which will contribute to a better understanding of public health dynamics.




## Questions:

In this project we will explore and answer the following questions;

 * How does a Bayesian modeling technique using MCMC improve knowledge on the rate of occurrence for some health indicators considering demographic conditions?

 * What are the temporal patterns or trends in health indicators throughout different years?
 
 * How do health indicators differ based on gender and race/ethnicity?
 
 * Can we discover clusters of areas with similar health indicators?
 
 * How can we manage missing information in the 'Gender' category?




```{r, eval=FALSE, echo=FALSE, include=FALSE}
# How does Bayesian inference differ from typical statistical methods in this context?

# What are the posterior distributions for the parameters of interest?
# How do the values of health indicators change with time?
# Have there been notable increases or declines in specific indicators?

# What is the difference in health indicator values between genders?
#  How do different racial and ethnic groupings compare on these indicators?
# How many natural clusters exist in the data when it comes to health indicators?
# What demographic and geographical variables distinguish these clusters?

# Do cyclical patterns or trends occur over time?
# How are these patterns related to external forces or events?
# Which imputation approaches can be used to estimate missing gender data?
# How does missing data affect the overall analysis and results?

```



## Dataset Description
 
The dataset consists of records of numerous health indicators that were gathered in many places over several years. A description of each variable is as follows;

 - Indicator Category: The broad category that a health indicator belongs under, such as cancer, substance abuse, and HIV/AIDS.
 
 - Indicator: The particular health metrics or results (e.g., Lung and heart cancer, Infant mortality rate, heart disease).
 
 - Year: The year when the health indicator was measured or the data was gathered.
 
 - Gender: The population group's gender (either male, female, or Both).
 
 - Race and Ethnicity: The population's race or ethnicity (e.g., White, Black, Hispanic).
 
 - Value: The dependent variable in our study, which is the health indicator's rate per 100,000 persons.
 
 - Place (Location): The geographical area (in this case large cities in the US) where the data was gathered.
 
 - BCHC Requested Methodology: The Big Cities Health Coalition's (BCHC) specific methods requested for gathering or analyzing the data.
 
 - Source: Where the data came from (e.g., Census, hospital records, National Center for Health Statistics (NCHS) and CDC reports).
 
 - Methods: The techniques employed to gather or calculate the values of the health indicators.
 
 - Notes: Extra details or observations on the data points.
 


## Statistical Methods

With "Value" as the dependent variable, we will use a number of sophisticated statistical computing techniques to answer the research questions and accomplish the project's goals. Methods like;

 * Exploratory Data Analysis
 
  - Descriptive Statistical Summary and Visualization (Histogram and Bar plots, Correlation Analysis)
  
 * Bayesian Model with Markov Chain Monte Carlo (MCMC): Utilize MCMC techniques to sample from the posterior distributions, such as Metropolis-Hastings and Gibbs sampling..
 
 * Longitudinal Analysis (Dynamics Across Year) : examines how the "Value" changes over time within the same entities (e.g., locations).
  
  
 * Model based Clustering: By using clustering analysis, data points are grouped so that they are more comparable to one another inside the same cluster than to those in other clusters.


 * EM Algorithm for missing Data: 
 
 
This project will allow to leverage a variety of statistical computing methods, offering a solid framework for analysis and valuable insights into some health indicator rate among large cities in the US.


## Load Data 

```{r,echo=FALSE, include=FALSE, eval=FALSE}

## Original values

Data_all <- read.csv("D:/WashU/First Year/Sem1/SDS5531_StatsComputing/Final_Project/Big_Cities_Health_Data_Inventory.csv")%>% mutate(Gender = as.factor(Gender), Race..Ethnicity = as.factor(Race..Ethnicity), Indicator.Category = as.factor(Indicator.Category), Indicator = as.factor(Indicator), Value = as.numeric(Value),Place = as.factor(Place))%>% dplyr::select(-BCHC.Requested.Methodology,-Source, -Methods, -Notes)

## Transformed Value

Data_all_trans <- read.csv("D:/WashU/First Year/Sem1/SDS5531_StatsComputing/Final_Project/Big_Cities_Health_Data_Inventory.csv")%>% mutate(Gender = as.factor(Gender), Race..Ethnicity = as.factor(Race..Ethnicity), Indicator.Category = as.factor(Indicator.Category), Indicator = as.factor(Indicator), Value = as.numeric(log(Value)),Place = as.factor(Place)) %>% dplyr::select(-BCHC.Requested.Methodology,-Source, -Methods, -Notes) 

## removing year ranges 

all_Data <- subset(Data_all, !grepl("-", Year))

all_Data_trans <- subset(Data_all_trans, !grepl("-", Year))
```



## Select by Health Indicator

```{r,echo=FALSE, include=FALSE, eval=FALSE}
Data <- read.csv("D:/WashU/First Year/Sem1/SDS5531_StatsComputing/Final_Project/Big_Cities_Health_Data_Inventory.csv") %>% filter(Indicator == "All Types of Cancer Mortality Rate (Age-Adjusted; Per 100,000 people)")%>% mutate(Gender = as.factor(Gender), Race..Ethnicity = as.factor(Race..Ethnicity), Indicator.Category = as.factor(Indicator.Category), Indicator = as.factor(Indicator), Year = as.numeric(Year), Value = as.numeric(Value),Place = as.factor(Place)) %>% dplyr::select(-BCHC.Requested.Methodology,-Source, -Methods, -Notes)



## Transformed

Data_trans <- read.csv("D:/WashU/First Year/Sem1/SDS5531_StatsComputing/Final_Project/Big_Cities_Health_Data_Inventory.csv") %>% filter(Indicator == "All Types of Cancer Mortality Rate (Age-Adjusted; Per 100,000 people)")%>% mutate(Gender = as.factor(Gender), Race..Ethnicity = as.factor(Race..Ethnicity), Indicator.Category = as.factor(Indicator.Category), Indicator = as.factor(Indicator), Year = as.numeric(Year), Value = as.numeric(log(Value)),Place = as.factor(Place)) %>% dplyr::select(-BCHC.Requested.Methodology,-Source, -Methods, -Notes)

## removing year ranges 

all_Data1 <- subset(Data, !grepl("-", Year))

all_Data1_trans <- subset(Data_trans, !grepl("-", Year))

```




```{r,echo=FALSE, include=FALSE, eval=FALSE}
Data1 <- read.csv("D:/WashU/First Year/Sem1/SDS5531_StatsComputing/Final_Project/Big_Cities_Health_Data_Inventory.csv") %>% filter(Indicator == "All-Cause Mortality Rate (Age-Adjusted; Per 100,000 people)")%>% mutate(Gender = as.factor(Gender), Race..Ethnicity = as.factor(Race..Ethnicity), Indicator.Category = as.factor(Indicator.Category), Indicator = as.factor(Indicator), Year = as.numeric(Year), Value = as.numeric(log(Value)),Place = as.factor(Place)) %>% dplyr::select(-BCHC.Requested.Methodology,-Source, -Methods, -Notes) 
```


```{r,echo=FALSE, include=FALSE, eval=FALSE}

# Histogram with density plot
fig1 <- ggplot(data = Data) +
  geom_histogram(aes(x = Value, y = ..density..), bins = 30, fill = 'skyblue', alpha = 0.5) +
  theme_bw()

fig2 <- ggplot(data = Data1) +
  geom_histogram(aes(x = Value, y = ..density..), bins = 30, fill = 'skyblue', alpha = 0.5) +
  theme_bw()

Data_all$Value <- log(Data_all$Value)

fig3 <- ggplot(data = Data_all) +
  geom_histogram(aes(x = Value, y = ..density..), bins = 100, fill = 'skyblue', alpha = 0.5) +
  theme_bw()


fig4 <- ggplot(data = Data_all, aes(x = Year, y = Value, color = Gender))+
  geom_point(alpha = 0.3)  + xlab("Year Starting from 2010")+
  ylab("Value") + ggtitle("Relationship Between Year and Response Value")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=8))


grid.arrange(fig1,fig2,fig3,fig4, nrow = 4)

```


```{r}
## Year and Value
fig5 <- ggplot(data = all_Data, aes(x = Year, y = Value, color = Gender))+
  geom_boxplot(alpha = 0.3)  + xlab("Year Starting from 2010")+
  ylab("Value") + ggtitle("Relationship Between Year and Response Value")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=8))

## adjust the year
```


## susetting all data by Health Indicator

```{r}
data_1 <- Data_all  %>% filter(Indicator.Category == "Infectious Disease")
data_2 <- Data_all  %>% filter(Indicator.Category == "Injury and Violence")
data_3 <- Data_all  %>% filter(Indicator.Category == "HIV/AIDS")
data_4 <- Data_all  %>% filter(Indicator.Category == "Nutrition, Physical Activity, & Obesity")

# mean(is.na(Data$Gender))

```




## Plots

```{r}
ggplot(data = data_3, aes(x = Year, y = Value, color = Gender))+
  geom_point(alpha = 0.3)  + xlab("Year Starting from 2010")+
  ylab("Value") + ggtitle("Relationship Between Year and Response Value")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=8))
```


## Exploring Trends

### Yearly trends

```{r}

# Aggregate the data to get mean Value per Year
yearly_trends <- all_Data1 %>%
  group_by(Year) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE))

# View the aggregated data
print(yearly_trends)


```


```{r}

# Plot mean Value over Years
ggplot(yearly_trends, aes(x = Year, y = mean_value)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  labs(title = "Temporal Trend of Health Indicator Values",
       x = "Year",
       y = "Mean Value (per 100,000 people)") +
  theme_minimal()


```


## Trend by Gender

```{r}
# Filter out missing Gender data
data_gender <- all_Data1 %>%
  filter(!is.na(Gender) & Gender != "")

# Aggregate data by Year and Gender
yearly_gender_trends <- data_gender %>%
  group_by(Year, Gender) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE))

# View the aggregated data
print(yearly_gender_trends)

```



```{r}
# Plot mean Value over Years by Gender
ggplot(yearly_gender_trends, aes(x = Year, y = mean_value, color = Gender, group = Gender)) +
  geom_line() +
  geom_point() +
  labs(title = "Temporal Trend of Health Indicator Values by Gender",
       x = "Year",
       y = "Mean Value (per 100,000 people)",
       color = "Gender") +
  theme_minimal()

```

## Trends by Race

```{r}
# Filter out missing Race/Ethnicity data
data_race <- all_Data1 %>%
  filter(!is.na(Race..Ethnicity) & Race..Ethnicity != "")

# Aggregate data by Year and Race/Ethnicity
yearly_race_trends <- data_race%>%
  group_by(Year, Race..Ethnicity) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE))

# View the aggregated data
print(yearly_race_trends)


```


```{r}
# Plot mean Value over Years by Race/Ethnicity
ggplot(yearly_race_trends, aes(x = Year, y = mean_value, color = Race..Ethnicity, group = Race..Ethnicity)) +
  geom_line() +
  geom_point() +
  labs(title = "Temporal Trend of Health Indicator Values by Race/Ethnicity",
       x = "Year",
       y = "Mean Value (per 100,000 people)",
       color = "Race/Ethnicity") +
  theme_minimal()

```

## Trends by Place

```{r}
# Aggregate data by Year and Place
yearly_place_trends <- all_Data1 %>%
  group_by(Year, Place) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE))

# View the aggregated data
print(yearly_place_trends)

```


```{r}
# Select top 5 Places with the most data points
top_places <- all_Data1 %>%
  group_by(Place) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  slice(1:5) %>%
  pull(Place)

# Filter data for top Places
top_places_data <- yearly_place_trends %>%
  filter(Place %in% top_places)

# Plot mean Value over Years for top Places
ggplot(top_places_data, aes(x = Year, y = mean_value, color = Place, group = Place)) +
  geom_line() +
  geom_point() +
  labs(title = "Temporal Trend of Health Indicator Values for Top Places",
       x = "Year",
       y = "Mean Value (per 100,000 people)",
       color = "Place") +
  theme_minimal()

```


```{r, echo=FALSE, include=FALSE, eval=FALSE}
table(Data_all$Indicator.Category)
```


This shows that analysis on the entire data will be challenging, so do in groups

## EM Algorithm For Missing Gender in certain Years

```{r}

## Load Data again

health_data <- read.csv("D:/WashU/First Year/Sem1/SDS5531_StatsComputing/Final_Project/Big_Health_data.csv", stringsAsFactors = FALSE) %>% dplyr::select(-BCHC.Requested.Methodology,-Source, -Methods, -Notes) 



# health indicator and years of interest
indicator_of_interest <- "All Types of Cancer Mortality Rate (Age-Adjusted; Per 100,000 people)"

years_of_interest <- c(2010, 2011, 2012, 2013, 2014,2015)

# Filter the dataset
data_subset <- health_data %>%
  filter(Indicator == indicator_of_interest,
         Year %in% years_of_interest)


# The number missing data in Gender Column
sum(is.na(data_subset$Gender) | data_subset$Gender == "")



```

```{r}

# Separate observed and missing Gender data
observed_data <- data_subset %>%
  filter(!is.na(Gender) & Gender != "")

## Latent
missing_data <- data_subset %>%
  filter(is.na(Gender) | Gender == "")

# Initialize parameters using observed data
initial_means <- observed_data %>%
  group_by(Gender) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE))

# Initial variance estimate
initial_variance <- var(observed_data$Value, na.rm = TRUE)

# Initialize parameters
mu_male <- initial_means$mean_value[initial_means$Gender == "Male"]
mu_female <- initial_means$mean_value[initial_means$Gender == "Female"]
mu_both <- initial_means$mean_value[initial_means$Gender == "Both"]
sigma2 <- initial_variance

```


## EM Loop 
```{r}
# Stopping Criterion parameters
epsilon <- 1e-6
max_iter <- 1000
log_likelihood_old <- -Inf



# Initialize iteration counter
iter <- 0
converged <- FALSE

# Store log-likelihoods 
log_likelihoods <- c()

# EM Algorithm
while (!converged && iter < max_iter) {
  iter <- iter + 1
  
  # E-step: Compute posterior probabilities for missing Gender
  # For each missing data point, calculate P(Gender = g | Y_i, current parameters)
  
  # Get current parameters
  params <- list(
    mu_male = mu_male,
    mu_female = mu_female,
    mu_both = mu_both,
    sigma2 = sigma2
  )
  
  # Calculate the likelihoods for each Gender category
  likelihoods <- sapply(c("Male", "Female", "Both"), function(g) {
    mu_g <- params[[paste0("mu_", tolower(g))]]
    dnorm(missing_data$Value, mean = mu_g, sd = sqrt(sigma2))
  })
  
  # Convert likelihoods to probabilities
  # Assuming equal prior probabilities for Genders
  posterior_probs <- likelihoods / rowSums(likelihoods)
  
  # M-step: Update parameters using observed data and expected counts from E-step
  
  # Combine observed data and missing data with estimated Gender probabilities
  # For observed data, the Gender is known
  observed_data$Gender_numeric <- observed_data$Gender
  observed_data$Male <- as.numeric(observed_data$Gender == "Male")
  observed_data$Female <- as.numeric(observed_data$Gender == "Female")
  observed_data$Both <- as.numeric(observed_data$Gender == "Both")
  
  # For missing data, use posterior probabilities
  missing_data$Male <- posterior_probs[, "Male"]
  missing_data$Female <- posterior_probs[, "Female"]
  missing_data$Both <- posterior_probs[, "Both"]
  
  # Combine datasets
  combined_data <- rbind(
    observed_data %>% select(Value, Male, Female, Both),
    missing_data %>% select(Value, Male, Female, Both)
  )
  
  # Update means
  mu_male_new <- sum(combined_data$Male * combined_data$Value) / sum(combined_data$Male)
  mu_female_new <- sum(combined_data$Female * combined_data$Value) / sum(combined_data$Female)
  mu_both_new <- sum(combined_data$Both * combined_data$Value) / sum(combined_data$Both)
  
  # Update variance
  residuals_squared <- combined_data$Male * (combined_data$Value - mu_male_new)^2 +
                       combined_data$Female * (combined_data$Value - mu_female_new)^2 +
                       combined_data$Both * (combined_data$Value - mu_both_new)^2
  sigma2_new <- sum(residuals_squared) / nrow(combined_data)
  
  # Compute log-likelihood
  log_likelihood_new <- sum(
    combined_data$Male * dnorm(combined_data$Value, mean = mu_male_new, sd = sqrt(sigma2_new), log = TRUE) +
    combined_data$Female * dnorm(combined_data$Value, mean = mu_female_new, sd = sqrt(sigma2_new), log = TRUE) +
    combined_data$Both * dnorm(combined_data$Value, mean = mu_both_new, sd = sqrt(sigma2_new), log = TRUE)
  )
  
  # Store log-likelihood for plotting
  log_likelihoods <- c(log_likelihoods, log_likelihood_new)
  
  # Check for convergence
  if (abs(log_likelihood_new - log_likelihood_old) < epsilon) {
    converged <- TRUE
  } else {
    # Update parameters for next iteration
    mu_male <- mu_male_new
    mu_female <- mu_female_new
    mu_both <- mu_both_new
    sigma2 <- sigma2_new
    log_likelihood_old <- log_likelihood_new
  }
  
  

  

}



## Plot of Convergence

plot(log_likelihoods, type = "o", col = "blue",
     xlab = "Iteration", ylab = "Log-likelihood",
     main = "EM Algorithm Convergence")


# Final parameter estimates

Missing_Gender_result <- data.frame(
  " mu_male" = mu_male_new,
  "mu_female" = mu_female_new,
  "mu_both" = mu_both_new,
  "sigma2" = sigma2_new,
  "log_like_old" = log_likelihood_new
)


kable(Missing_Gender_result, digits = 8, caption = "Estimated Parameters")%>%
  kable_styling(position = "center",
              latex_options = "HOLD_position")


## Plugging in estimate into data

# For missing data, assign Gender based on highest posterior probability
missing_data$Gender_estimated <- apply(posterior_probs, 1, function(x) {
  genders <- c("Male", "Female", "Both")
  genders[which.max(x)]
})

# Combine observed and missing data
final_data <- rbind(
  observed_data %>% select(-Male, -Female, -Both, -Gender_numeric),
  missing_data %>% select(-Male, -Female, -Both, - Gender_estimated)
)

# # Update the Gender column for missing data
# final_data$Gender[is.na(final_data$Gender) | final_data$Gender == ""] <-
#   final_data$Gender_estimated[is.na(final_data$Gender) | final_data$Gender == ""]

# View the updated data
head(final_data)


## write out algorithm in words

```




# Clustering
```{r, echo=FALSE, include=FALSE, eval=FALSE}
# K-means clustering

kmeans_result <- kmeans(Data_all$Value, centers = 100)
data$cluster <- as.factor(kmeans_result$cluster)

# Visualize clusters
ggplot(data, aes(x = var1, y = var2, color = cluster)) +
  geom_point() +
  theme_minimal()

```

