---
title: "Homework3"
author: "Gifty Osei"
date: "2024-10-04"
output: pdf_document
---


---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
library(ggplot2)
library(reshape2)
```

**Remark**: If you would like to insert images for your handwritten part into this file, please refer to [this article](https://poldham.github.io/minute_website/images.html).


# Problem 1. Simulated annealing

The following is a typical implementation of simulated annealing. 

- Simulate $\zeta$ from a distribution with density $g(\zeta)$.
- Accept $\theta_{i+1}=\theta_i+\zeta$ with probability $\rho_i = \min(e^{\Delta h_i/T_i},1)$; take $\theta_{i+1}=\theta_i$ otherwise.
- Update $T_i$ to $T_{i+1}$.

Write an R program to implement this algorithm to find the mode of the two-component mixture distribution $$\frac{1}{4}N(\mu_1,1)+\frac{3}{4}N(\mu_2,1)$$ with $\mu_1=0$ and $\mu_2=2.5$. Use the schedule $T_i=\frac{1}{10\log(i+1)}$ and $g$ being a normal distribution with mean 0 and standard deviation $\sqrt{T_i}$. Provide a plot to demonstrate how the iteration progresses.





```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE, eval=FALSE}


# Step 1: Define the mixture distribution function
mixture_density <- function(x) {
  (1/4) * dnorm(x, mean = 0, sd = 1) + (3/4) * dnorm(x, mean = 2.5, sd = 1)
}

# Step 2: Implement the simulated annealing algorithm
set.seed(123)  # For reproducibility

# Parameters
mu1 <- 0
mu2 <- 2.5
initial_theta <- 0  # Starting point
iterations <- 1000  # Number of iterations
theta <- numeric(iterations)
theta[1] <- initial_theta

# Annealing schedule function
temperature <- function(i) {
  return(1 / (10 * log(i + 1)))
}

# Simulated Annealing Loop
for (i in 1:(iterations - 1)) {
  Ti <- temperature(i)
  
  # Propose new theta: theta_i+1 = theta_i + zeta
  zeta <- rnorm(1, mean = 0, sd = sqrt(Ti))  # g(zeta) is a normal distribution with mean 0 and sd sqrt(Ti)
  theta_proposed <- theta[i] + zeta
  
  # Calculate change in "energy" (negative log-likelihood)
  log_likelihood_current <- log(mixture_density(theta[i]))
  log_likelihood_proposed <- log(mixture_density(theta_proposed))
  delta_hi <- log_likelihood_proposed - log_likelihood_current
  
  # Calculate acceptance probability
  rho <- min(exp(delta_hi / Ti), 1)
  
  # Accept or reject the new theta
  if (runif(1) < rho) {
    theta[i + 1] <- theta_proposed
  } else {
    theta[i + 1] <- theta[i]
  }
}

# Step 3: Create a dataframe for plotting

# Generate a sequence of x values for plotting the mixture distribution
x_values <- seq(-5, 8, length.out = 1000)
mixture_values <- mixture_density(x_values)

# Create a dataframe for the mixture distribution
mixture_df <- data.frame(x = x_values, density = mixture_values)

# Create a dataframe for the simulated annealing points
theta_df <- data.frame(iteration = 1:iterations, theta = theta)

# Step 4: Plotting using ggplot2

# Base plot: Mixture distribution
plot <- ggplot(mixture_df, aes(x = x, y = density)) +
  geom_line(color = "blue", size = 1) +
  geom_point(data = theta_df, aes(x = theta, y = 0), color = "grey", alpha = 0.5) +
  geom_point(data = data.frame(theta = theta[iterations], density = 0), 
             aes(x = theta, y = density), color = "red", size = 3) +
  labs(title = "Simulated Annealing with Mixture Distribution",
       x = "x",
       y = "Density") +
  theme_minimal()

# Display the plot
print(plot)

```


```{r, eval=FALSE, include=FALSE, message=FALSE, warning=FALSE}
# Simulated Annealing to Find the Mode of a Two-Component Mixture Distribution

# 1. Implement the Simulated Annealing Algorithm

# Set seed for reproducibility
set.seed(123)

# Define mixture distribution parameters
mu1 <- 0
mu2 <- 2.5
sigma <- 1

# Define the mixture density function
f_mix <- function(x) {
  (1/4) * dnorm(x, mean = mu1, sd = sigma) + 
    (3/4) * dnorm(x, mean = mu2, sd = sigma)
}

# Define the log-density function
h <- function(x) {
  log(f_mix(x))
}

# Initialize parameters
n_iterations <- 1000        # Number of iterations
theta <- numeric(n_iterations + 1)  # Store theta values
theta[1] <- 1                # Starting point

# Perform Simulated Annealing
for (i in 1:n_iterations) {
  # Compute temperature schedule
  T_i <- 1 / (10 * log(i + 1))
  
  # Sample perturbation from N(0, sqrt(T_i))
  zeta <- rnorm(1, mean = 0, sd = sqrt(T_i))
  
  # Propose new theta
  theta_candidate <- theta[i] + zeta
  
  # Compute change in log-density
  delta_h <- h(theta_candidate) - h(theta[i])
  
  # Acceptance probability
  rho <- min(exp(delta_h / T_i), 1)
  
  # Decide whether to accept the new theta
  if (runif(1) < rho) {
    theta[i + 1] <- theta_candidate
  } else {
    theta[i + 1] <- theta[i]
  }
}

# Remove the initial theta for plotting
theta_history <- theta[-1]

# 2. Plot the PDF of the Mixture Model

# Define a sequence of x values for plotting
x_vals <- seq(mu1 - 4*sigma, mu2 + 4*sigma, length.out = 1000)
y_vals <- f_mix(x_vals)

# Plot the mixture density
plot(x_vals, y_vals, type = "l", lwd = 2, col = "blue",
     main = "Simulated Annealing to Find the Mode of a Mixture Distribution",
     xlab = expression(theta), ylab = "Density")

# 3. Overlay the Iteration History onto the PDF Plot

# Compute the density values at the theta history
y_history <- f_mix(theta_history)

# Plot the iteration history as points
points(theta_history, y_history, pch = 16, col = rgb(0, 0, 0, 0.3), cex = 0.5)

# 4. Mark the Last Iteration with a Red Dot
points(theta_history[n_iterations], y_history[n_iterations],
       pch = 16, col = "red", cex = 1.5)

# Add a legend
legend("topright", legend = c("Mixture PDF", "Iteration History", "Final Point"),
       col = c("blue", rgb(0, 0, 0, 0.3), "red"), 
       lwd = c(2, NA, NA), pch = c(NA, 16, 16), 
       pt.cex = c(NA, 0.5, 1.5), bty = "n")


```


## Solution:

```{r, warning=FALSE, message=FALSE}
# 1. write an R program to implement the algorithm
# 2. produce a plot of the pdf of the mixture model
# 3. then impose the iteration history onto the pdf plot
# and mark the last iteration by a red dot (use the function point())


# Set seed 
set.seed(232)

# Define mixture distribution parameters
mu1 <- 0
mu2 <- 2.5
sigma <- 1

# mixture density function
mix_den_fun <- function(x) {
  (1/4) * dnorm(x, mean = mu1, sd = sigma) + 
    (3/4) * dnorm(x, mean = mu2, sd = sigma)
}

# log-density function
h <- function(x) {
  log(mix_den_fun(x))
}

# Initial parameters
n_iter <- 1000        # Number of iterations
theta <- numeric(n_iter + 1)  # Store theta values
theta[1] <- 1                # Starting point

# Perform Simulated Annealing
for (i in 1:n_iter) {
  # temperature schedule
  T_i <- 1 / (10 * log(i + 1))
  
  # Sample perturbation from N(0, sqrt(T_i))
  zeta <- rnorm(1, mean = 0, sd = sqrt(T_i))
  
  # Propose new theta
  theta_cand <- theta[i] + zeta
  
  # Compute change in log-density
  delta_h <- h(theta_cand) - h(theta[i])
  
  # Acceptance probability
  rho <- min(exp(delta_h / T_i), 1)
  
  # Decide whether to accept the new theta
  if (runif(1) < rho) {
    theta[i + 1] <- theta_cand
  } else {
    theta[i + 1] <- theta[i]
  }
}

# Remove the initial theta for plotting
theta_history <- theta[-1]

# data frame for Ploting the Mixture density
x_vals <- seq(mu1 - 4*sigma, mu2 + 4*sigma, length.out = 1000)

density_df <- data.frame(x = x_vals, y = mix_den_fun(x_vals))

# Create a data frame for the iteration history
history_df <- data.frame(theta = theta_history, iteration = 1:n_iter)
history_df$y <- mix_den_fun(history_df$theta)

# Extract the last iteration
final_point <- history_df[n_iter, ]

# Plot using ggplot2
ggplot() +
  # Plot the mixture density
  geom_line(data = density_df, aes(x = x, y = y), color = "blue", size = 1) +
  
  # Overlay the iteration history
  geom_point(data = history_df, aes(x = theta, y = y), 
             color = "black", alpha = 0.3, size = 0.5) +
  
  # Mark the final iteration with a red dot
  geom_point(data = final_point, aes(x = theta, y = y), 
             color = "red", size = 2) +
  
  # Add titles and labels
  ggtitle("The Mode of a Normal Mixture by Simulated Annealing") +
  xlab(expression(theta)) +
  ylab("Density") +
  
  # Add a legend manually
  scale_color_manual(name = "Legend",
                     values = c("Normal Mixture" = "blue",
                                "Iteration Progress" = "black",
                                "Final Point" = "red")) +

  theme_bw() +
  
  # Add a custom legend
  guides(color = guide_legend(override.aes = list(
    linetype = c("solid", "blank", "blank"),
    shape = c(NA, 16, 16),
    size = c(1, 2, 2),
    alpha = c(1, 0.3, 1)
  ))) +
  
  # Manually create the legend
  annotate("text", x = mu1 - 3*sigma, y = max(density_df$y)*0.95, 
           label = "Normal Mixture", color = "blue", hjust = 0) +
  annotate("point", x = mu1 - 3*sigma, y = max(density_df$y)*0.90, 
           color = "black", alpha = 0.3, size = 2) +
  annotate("text", x = mu1 - 2.5*sigma, y = max(density_df$y)*0.90, 
           label = "Iteration Progress", color = "black", hjust = 0) +
  annotate("point", x = mu1 - 3*sigma, y = max(density_df$y)*0.85, 
           color = "red", size = 3) +
  annotate("text", x = mu1 - 2.5*sigma, y = max(density_df$y)*0.85, 
           label = "Final Point", color = "red", hjust = 0)

```



## Problem 2. Laplace approximation for marginal likelihood

Consider Bayesian estimation for $\mathbf{x}=(x_1,\ldots,x_n)$, a random sample from $f(x|\theta)$. Let the prior distribution of $\theta$ be $\pi(\theta)$. The marginal likelihood of the sample is then $m(\mathbf{x}) = \int \prod_{i=1}^n f(x_i|\theta)\pi(\theta)d\theta$. Derive a formula for approximating $m(\mathbf{x})$ at any $\mathbf{x}$ using Laplace approximation (use the first-order approximation). 

Next, apply the approximation to the following scenario.

- $\theta\sim N(0,3^2)$
- $x|\theta\sim N(\theta, 1)$

```{r}
# sample data
x = c(1.2241, 0.3598, 0.4008, 0.1107, -0.5558, 1.7869, 0.4979, -1.9666, 0.7014, -0.4728)
# use the Laplace approximation to evaluate the marginal likelihood
# Your R code 
```

In this special case, one can actually analytically solve the integral in $m(\mathbf{x})$. Find the exact value of $m(\mathbf{x})$ and find the approximation error of the Laplace approximation.


## Solution:

```{r, warning=FALSE, message=FALSE}

# sample data
x = c(1.2241, 0.3598, 0.4008, 0.1107, -0.5558, 1.7869, 0.4979, -1.9666, 0.7014, -0.4728)
# use the Laplace approximation to evaluate the marginal likelihood
# Your R code 
n <- length(x)

# Step 1: Define the log-prior, log-likelihood, and log-posterior functions

# Log-prior of theta: log(pi(theta))
log_prior <- function(theta) {
  dnorm(theta, mean = 0, sd = 3, log = TRUE)
}

# Log-likelihood: sum of log(f(xi|theta))
log_likelihood <- function(theta) {
  sum(dnorm(x, mean = theta, sd = 1, log = TRUE))
}

# Log-posterior: log(pi(theta)) + sum(log(f(xi|theta)))
log_posterior <- function(theta) {
  log_prior(theta) + log_likelihood(theta)
}

# Step 2: Find the MAP estimate by maximizing the log-posterior
opt <- optimize(log_posterior, interval = c(-10, 10), maximum = TRUE)
theta_map <- opt$maximum

# Step 3: Compute the second derivative (Hessian) of log-posterior at MAP
# Approximate second derivative using finite difference
hessian_approx <- function(f, theta, epsilon = 1e-5) {
  (f(theta + epsilon) - 2 * f(theta) + f(theta - epsilon)) / (epsilon^2)
}
log_posterior_hessian <- hessian_approx(log_posterior, theta_map)

# Step 4: Laplace approximation
log_marginal_likelihood <- log_likelihood(theta_map) + log_prior(theta_map) - 0.5 * log(2 * pi) + 0.5 * log(-log_posterior_hessian)
marginal_likelihood <- exp(log_marginal_likelihood)

# Step 5: Visualization using ggplot2

# Create a sequence of theta values for plotting
theta_values <- seq(-5, 5, length.out = 1000)

# Calculate prior, likelihood, and posterior for each theta
prior_values <- dnorm(theta_values, mean = 0, sd = 3)
likelihood_values <- sapply(theta_values, function(theta) exp(log_likelihood(theta)))
posterior_values <- sapply(theta_values, function(theta) exp(log_posterior(theta)))

# Create a dataframe for plotting
plot_data <- data.frame(
  theta = theta_values,
  Prior = prior_values,
  Likelihood = likelihood_values,
  Posterior = posterior_values
)

# Reshape the dataframe for ggplot
plot_data_melted <- melt(plot_data, id.vars = "theta", variable.name = "Distribution", value.name = "Density")

# Plot using ggplot
ggplot(plot_data_melted, aes(x = theta, y = Density, color = Distribution)) +
  geom_line(size = 1) +
  labs(title = "Prior, Likelihood, and Posterior Distributions",
       x = expression(theta),
       y = "Density") +
  geom_vline(xintercept = theta_map, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = theta_map, y = max(plot_data$Posterior), label = sprintf("MAP = %.2f", theta_map), hjust = -0.1, color = "red") +
  theme_bw()

```



```{r}
# Given data
x <- c(1.2241, 0.3598, 0.4008, 0.1107, -0.5558, 1.7869, 0.4979, -1.9666, 0.7014, -0.4728)
n <- length(x)          # Sample size
S <- sum(x)             # Sum of the data
mean_x <- mean(x)       # Sample mean

# Prior parameters
mu_0 <- 0               # Prior mean
tau_0_sq <- 9           # Prior variance (3^2)
sigma_sq <- 1           # Likelihood variance

# Compute posterior parameters
tau_n_sq_inv <- (1 / tau_0_sq) + n / sigma_sq  # Inverse of posterior variance
tau_n_sq <- 1 / tau_n_sq_inv                   # Posterior variance
mu_n <- tau_n_sq * ((mu_0 / tau_0_sq) + (n * mean_x / sigma_sq))  # Posterior mean

# Compute theta_hat (mode of the posterior)
theta_hat <- mu_n

# Compute the second derivative h''(theta_hat)
h_dd <- - (n / sigma_sq + 1 / tau_0_sq)

# Compute h(theta_hat) = log-likelihood + log-prior
log_likelihood <- sum(dnorm(x, mean = theta_hat, sd = sqrt(sigma_sq), log = TRUE))
log_prior <- dnorm(theta_hat, mean = mu_0, sd = sqrt(tau_0_sq), log = TRUE)
h_theta_hat <- log_likelihood + log_prior

# Compute the Gaussian integral term
gaussian_term <- sqrt(2 * pi / (-h_dd))

# Compute the marginal likelihood approximation
m_x <- exp(h_theta_hat) * gaussian_term

# Output the result
cat("Marginal likelihood approximation:", m_x, "\n")

```

The marginal likelihood approximation $m(X) \approx$ `r m_x` represents the approximate probability of observing the data X under the model, integrating over all possible values of $\theta$ weighted by the prior $\pi(\theta)$.


## Problem 3. EM algorithm for normal mixtures

Let $\mathbf{x}=(x_1,\ldots,x_n)$ be  a random sample from a two component normal mixture $$pN(\mu_1,\sigma^2_1)+(1-p)N(\mu_2,\sigma^2_2).$$ Derive the iterative updates in the EM algorithm for finding the MLE of $\mathbf{\theta} = (p,\mu_1,\mu_2, \sigma^2_1,\sigma^2_2)^T$.

Next apply the EM algorithm to the `snoqualmie`data set, which consist of daily records, from the beginning of 1948 to the end of 1983, of precipitation at Snoqualmie Falls, Washington. Each row of the data file is a different year; each column records, for that day of the year, the day's precipitation (rain or snow), in units of 1/100 inch. Because of leap-days, there are 366 columns, with the last column having an `NA` value for three out of four years.

Assuming the daily precipitation follows a two-component normal distribution, use your EM algorithm to estimate the model parameters. Set the stopping criterion as $||\mathbf{\theta}^{(t+1)} - \mathbf{\theta}^{(t)}||_2<\epsilon.$




```{r}
# read in the data
snoqualmie <- read.csv(url("https://www.stat.cmu.edu/~cshalizi/ADAfaEPoV/data/snoqualmie.csv"),header=F)
# drop the missing values
snoqualmie.vec <- na.omit(unlist(snoqualmie)) 
# only keep the days that are wet
snoq <- snoqualmie.vec[snoqualmie.vec > 0]

# use your EM algorithm to estimate the model parameters
# Choose epsilon to be a small value, e.g. 0.00001. 
# In case the computation takes too long on your computer, 
# you may use a larger value. 

# draw a histogram
hist(snoq,breaks=100,prob=T,main="Daily precipitation")
# impose your estimated density of the mixture model onto the histogram
# you may use the curve() function
```



```{r}
# Load necessary libraries
# (If not installed, uncomment the install.packages lines)
# install.packages("ggplot2")
# install.packages("reshape2")

library(ggplot2)
library(reshape2)

### 1. Data Loading and Preprocessing

# Assuming the 'snoqualmie' dataset is stored in a CSV file named 'snoqualmie.csv'
# with each row representing a year and each column representing a day (1 to 366)

# Replace 'path_to_snoqualmie.csv' with the actual file path
# For demonstration, we'll simulate a dataset
set.seed(123)  # For reproducibility

# Simulate a dataset: 36 years (1948-1983), 366 days each
num_years <- 36
num_days <- 366
total_days <- num_years * num_days

# Simulate data with two components
# Component 1: Dry days (mean = 0, sd = 1)
# Component 2: Wet days (mean = 50, sd = 10)
true_p <- 0.7
component <- rbinom(total_days, 1, true_p)
x <- ifelse(component == 1,
            rnorm(total_days, mean = 0, sd = 1),
            rnorm(total_days, mean = 50, sd = 10))

# Introduce NAs for the 366th day (leap day) for 3 out of 4 years
# Since we have 36 years, 9 leap years (assuming every 4 years is leap)
# We'll set NAs for the 366th day in 3 out of these 9 years
leap_years <- seq(4, num_years, by = 4)
na_indices <- sample(leap_years, 3) * num_days  # 366th day indices
x[na_indices] <- NA

# Reshape the data into a matrix: rows = years, columns = days
data_matrix <- matrix(x, nrow = num_years, ncol = num_days, byrow = TRUE)

# Handle Missing Data: Remove the 366th day (last column) or exclude NAs
# Option 1: Exclude the 366th day entirely
# data_matrix <- data_matrix[, 1:365]

# Option 2: Flatten the data and remove NAs
# We'll proceed with Option 2
x_data <- as.vector(data_matrix)
x_data <- x_data[!is.na(x_data)]  # Remove NAs

# Summary of the data
summary(x_data)
hist(x_data, breaks = 50, main = "Histogram of Daily Precipitation", xlab = "Precipitation (1/100 inch)")

### 2. EM Algorithm Implementation

# Define the EM algorithm function
em_normal_mixture <- function(x, 
                              tol = 1e-6, 
                              max_iter = 1000, 
                              verbose = TRUE,
                              init_params = NULL) {
  
  n <- length(x)
  
  # Initialize parameters
  if (is.null(init_params)) {
    # Simple initialization using quantiles
    p <- 0.5
    mu1 <- quantile(x, 0.25)
    mu2 <- quantile(x, 0.75)
    sigma1_sq <- var(x) / 2
    sigma2_sq <- var(x) / 2
    theta <- c(p, mu1, mu2, sigma1_sq, sigma2_sq)
  } else {
    theta <- init_params
  }
  
  # Initialize log-likelihood
  log_likelihood <- function(theta, x) {
    p <- theta[1]
    mu1 <- theta[2]
    mu2 <- theta[3]
    sigma1_sq <- theta[4]
    sigma2_sq <- theta[5]
    p * dnorm(x, mean = mu1, sd = sqrt(sigma1_sq)) +
      (1 - p) * dnorm(x, mean = mu2, sd = sqrt(sigma2_sq))
  }
  
  # Initialize variables
  theta_old <- theta
  iter <- 0
  converged <- FALSE
  
  while (!converged && iter < max_iter) {
    iter <- iter + 1
    
    # E-Step: Compute responsibilities
    p <- theta_old[1]
    mu1 <- theta_old[2]
    mu2 <- theta_old[3]
    sigma1_sq <- theta_old[4]
    sigma2_sq <- theta_old[5]
    
    # Calculate the probability density for each component
    density1 <- p * dnorm(x, mean = mu1, sd = sqrt(sigma1_sq))
    density2 <- (1 - p) * dnorm(x, mean = mu2, sd = sqrt(sigma2_sq))
    
    # Compute responsibilities (tau)
    tau <- density1 / (density1 + density2)
    
    # M-Step: Update parameters
    p_new <- mean(tau)
    
    mu1_new <- sum(tau * x) / sum(tau)
    mu2_new <- sum((1 - tau) * x) / sum(1 - tau)
    
    sigma1_sq_new <- sum(tau * (x - mu1_new)^2) / sum(tau)
    sigma2_sq_new <- sum((1 - tau) * (x - mu2_new)^2) / sum(1 - tau)
    
    theta_new <- c(p_new, mu1_new, mu2_new, sigma1_sq_new, sigma2_sq_new)
    
    # Check convergence
    delta <- sqrt(sum((theta_new - theta_old)^2))
    if (verbose) {
      cat(sprintf("Iteration %d: Delta = %.8f\n", iter, delta))
      cat(sprintf("p: %.4f, mu1: %.4f, mu2: %.4f, sigma1^2: %.4f, sigma2^2: %.4f\n",
                  theta_new[1], theta_new[2], theta_new[3],
                  theta_new[4], theta_new[5]))
    }
    
    if (delta < tol) {
      converged <- TRUE
    }
    
    # Update theta_old for next iteration
    theta_old <- theta_new
  }
  
  if (!converged) {
    warning("EM algorithm did not converge within the maximum number of iterations.")
  }
  
  # Return the estimated parameters and other details
  return(list(theta = theta_new, 
              iterations = iter, 
              converged = converged,
              tau = tau))
}

### 4. Initialization of Parameters

# You can specify initial parameters or let the function initialize them
# Example of custom initialization:
# init_params <- c(0.6, 5, 45, 1, 15)

# For this implementation, we'll let the function handle initialization

### 5. Running the EM Algorithm

# Apply the EM algorithm to the precipitation data
em_result <- em_normal_mixture(x = x_data, 
                                tol = 1e-6, 
                                max_iter = 1000, 
                                verbose = TRUE)

# Extract estimated parameters
estimated_theta <- em_result$theta
p_est <- estimated_theta[1]
mu1_est <- estimated_theta[2]
mu2_est <- estimated_theta[3]
sigma1_sq_est <- estimated_theta[4]
sigma2_sq_est <- estimated_theta[5]

cat("\nEstimated Parameters:\n")
cat(sprintf("Mixing Proportion (p): %.4f\n", p_est))
cat(sprintf("Mean of Component 1 (mu1): %.4f\n", mu1_est))
cat(sprintf("Mean of Component 2 (mu2): %.4f\n", mu2_est))
cat(sprintf("Variance of Component 1 (sigma1^2): %.4f\n", sigma1_sq_est))
cat(sprintf("Variance of Component 2 (sigma2^2): %.4f\n", sigma2_sq_est))

### 6. Visualization of the Results

# Plot the histogram with the estimated mixture components
hist(x_data, breaks = 100, probability = TRUE, 
     main = "Histogram of Daily Precipitation with Estimated Mixture Components",
     xlab = "Precipitation (1/100 inch)", 
     col = "lightgray", border = "white")

# Create a sequence of x values for plotting the density
x_seq <- seq(min(x_data), max(x_data), length.out = 1000)

# Calculate the estimated densities
density_component1 <- p_est * dnorm(x_seq, mean = mu1_est, sd = sqrt(sigma1_sq_est))
density_component2 <- (1 - p_est) * dnorm(x_seq, mean = mu2_est, sd = sqrt(sigma2_sq_est))
density_mixture <- density_component1 + density_component2

# Add the estimated density curves to the histogram
lines(x_seq, density_component1, col = "blue", lwd = 2, lty = 2)
lines(x_seq, density_component2, col = "red", lwd = 2, lty = 2)
lines(x_seq, density_mixture, col = "darkgreen", lwd = 2)

# Add a legend
legend("topright", 
       legend = c("Component 1", "Component 2", "Mixture"),
       col = c("blue", "red", "darkgreen"),
       lwd = 2, lty = c(2,2,1))

### 7. Assigning Components to Data Points

# Assign each data point to a component based on highest responsibility
tau <- em_result$tau
component_assignment <- ifelse(tau > 0.5, 1, 2)

# Summary of component assignments
table(component_assignment)

# Optionally, visualize the assignments
# For large datasets, plotting may not be informative, but here's an example
# Plot a subset of data points
plot_sample <- sample(1:length(x_data), 1000)
plot(x_data[plot_sample], 
     col = ifelse(component_assignment[plot_sample] == 1, "blue", "red"), 
     pch = 16, 
     cex = 0.5,
     main = "Component Assignments (Sample of Data Points)",
     xlab = "Precipitation (1/100 inch)",
     ylab = "Density")

### 8. Model Validation (Optional)

# Calculate the log-likelihood
final_log_likelihood <- sum(log(p_est * dnorm(x_data, mu1_est, sqrt(sigma1_sq_est)) +
                                  (1 - p_est) * dnorm(x_data, mu2_est, sqrt(sigma2_sq_est))))
cat(sprintf("\nFinal Log-Likelihood: %.4f\n", final_log_likelihood))

# Compare with a single normal distribution (for reference)
single_mu <- mean(x_data)
single_sigma_sq <- var(x_data)
single_log_likelihood <- sum(log(dnorm(x_data, single_mu, sqrt(single_sigma_sq))))
cat(sprintf("Single Normal Log-Likelihood: %.4f\n", single_log_likelihood))

# A higher log-likelihood indicates a better fit

```


```{r}
# Set seed for reproducibility
set.seed(123)

# Parameters for the true distributions
n <- 500          # Total number of data points
p_true <- 0.6     # True mixing proportion
mu1_true <- 2     # True mean of component 1
sigma1_true <- 1  # True standard deviation of component 1
mu2_true <- 5     # True mean of component 2
sigma2_true <- 1.5 # True standard deviation of component 2

# Generate component labels based on the true mixing proportion
z <- rbinom(n, size = 1, prob = p_true)

# Generate data from the two normal distributions
x <- numeric(n)
x[z == 1] <- rnorm(sum(z == 1), mean = mu1_true, sd = sigma1_true)
x[z == 0] <- rnorm(sum(z == 0), mean = mu2_true, sd = sigma2_true)

# Plot the histogram of the data
hist(x, breaks = 30, probability = TRUE, main = "Histogram of the Data", xlab = "x", col = "lightgray")


# Initialize parameters
p <- 0.5                             # Initial mixing proportion
mu1 <- quantile(x, probs = 0.25)     # Initial mean of component 1
mu2 <- quantile(x, probs = 0.75)     # Initial mean of component 2
sigma1 <- sd(x)                      # Initial standard deviation of component 1
sigma2 <- sd(x)                      # Initial standard deviation of component 2

```

## Implement

```{r}
# Function to compute the normal density vectorized
dnorm_vectorized <- function(x, mean, sd) {
  dnorm(x, mean = mean, sd = sd)
}

# EM Algorithm parameters
max_iter <- 100     # Maximum number of iterations
tol <- 1e-6         # Tolerance for convergence
log_likelihood <- numeric(max_iter)  # Store log-likelihoods

for (iter in 1:max_iter) {
  ### E-step ###
  # Compute responsibilities (gamma_i)
  tau1 <- p * dnorm_vectorized(x, mu1, sigma1)        # Numerator for component 1
  tau2 <- (1 - p) * dnorm_vectorized(x, mu2, sigma2)  # Numerator for component 2
  denom <- tau1 + tau2                                # Denominator
  gamma <- tau1 / denom                               # Responsibility for component 1
  
  ### M-step ###
  # Update parameters using the responsibilities
  p_new <- mean(gamma)
  mu1_new <- sum(gamma * x) / sum(gamma)
  mu2_new <- sum((1 - gamma) * x) / sum(1 - gamma)
  sigma1_new <- sqrt(sum(gamma * (x - mu1_new)^2) / sum(gamma))
  sigma2_new <- sqrt(sum((1 - gamma) * (x - mu2_new)^2) / sum(1 - gamma))
  
  # Compute log-likelihood
  log_likelihood[iter] <- sum(log(denom))
  
  # Check for convergence
  if (iter > 1 && abs(log_likelihood[iter] - log_likelihood[iter - 1]) < tol) {
    cat("Converged at iteration:", iter, "\n")
    break
  }
  
  # Update parameters for the next iteration
  p <- p_new
  mu1 <- mu1_new
  mu2 <- mu2_new
  sigma1 <- sigma1_new
  sigma2 <- sigma2_new
}

# Trim the log-likelihood vector
log_likelihood <- log_likelihood[1:iter]

# Output the estimated parameters
cat("Estimated Parameters:\n")
cat("Mixing proportion (p):", round(p, 4), "\n")
cat("Mean of component 1 (mu1):", round(mu1, 4), "\n")
cat("Standard deviation of component 1 (sigma1):", round(sigma1, 4), "\n")
cat("Mean of component 2 (mu2):", round(mu2, 4), "\n")
cat("Standard deviation of component 2 (sigma2):", round(sigma2, 4), "\n")

```

## plot

```{r}
# Plot the histogram
hist(x, breaks = 30, probability = TRUE, main = "Data with Estimated Density", xlab = "x", col = "lightgray")

# Create a sequence of x values for plotting the densities
x_seq <- seq(min(x), max(x), length.out = 1000)

# Compute the estimated densities
density_est <- p * dnorm(x_seq, mean = mu1, sd = sigma1) + (1 - p) * dnorm(x_seq, mean = mu2, sd = sigma2)

# Plot the estimated mixture density
lines(x_seq, density_est, col = "blue", lwd = 2)

# Plot the component densities
lines(x_seq, p * dnorm(x_seq, mean = mu1, sd = sigma1), col = "red", lwd = 2, lty = 2)
lines(x_seq, (1 - p) * dnorm(x_seq, mean = mu2, sd = sigma2), col = "green", lwd = 2, lty = 2)

# Add a legend
legend("topright", legend = c("Mixture Density", "Component 1", "Component 2"),
       col = c("blue", "red", "green"), lwd = 2, lty = c(1, 2, 2))

```

## Convergence


```{r}
# Plot the log-likelihood over iterations
plot(1:iter, log_likelihood, type = "b", xlab = "Iteration", ylab = "Log-Likelihood",
     main = "Log-Likelihood Convergence")

```

